---
title: "Proposal: Tracing COVID-19 Arrival in Europe: Air Travel, Epidemic Seeding, and Excess Mortality"
author:
  - name: Maximilian Elixhauser
    corresponding: true
    email: maximilian.elixhauser@stud.plus.ac.at
    roles:
      - Author
    affiliations:
      - University of Salzburg
keywords:
  - Global Mobility Network
  - COVID-19
  - Excess Mortality
  - SIR Model
  - Report
date: last-modified
bibliography: proposal_ref.bib
number-sections: true
format:
  html:
    toc: true
    toc-depth: 3
---

## Introduction

In January 2020, a single infected traveler on a long-haul flight could spark an outbreak thousands of kilometers away—turning global aviation into a high-speed conveyor belt for COVID-19. Indeed, early analyses demonstrated that aviation can rapidly reduce effective distances between cities, accelerating worldwide transmission [@chinazzi2020; @sun2022_a]. At the same time, countries exhibited **large variability in excess mortality**, often shaped by **demographic** (e.g., population age, life expectancy) and **economic** (e.g., GDP, healthcare capacity) factors. For instance, @rahmanian2024 identified government effectiveness, vaccination coverage, and population age structure as key drivers of cross-national differences in pandemic mortality across 21 countries. Similarly, @matveeva2023 noted that lower GDP per capita and slower vaccination rates were correlated with higher excess mortality among European nations, reflecting a synergistic effect of vaccination coverage and broader healthcare resources on reducing COVID-19–associated deaths.

A less investigated aspect, however, is **epidemic arrival time**: Did countries “hit first” by the virus—when non-pharmaceutical interventions (NPI) were still immature—face worse overall mortality outcomes? This project aims to explore how early seeding of COVID-19 might have compounded the effects of the aforementioned demographic and economic predictors, ultimately shaping cross-national excess mortality patterns.

As outlined in the Z_GIS Thesis Blog by @neuwirth2024, the **hypothesis** states that **locations hit early and unprepared** faced more severe outbreaks, characterized by:

- **Longer durations** without access to vaccination  
- **Lack of effective non-pharmaceutical interventions (NPIs)** during early waves  
- **Additional outbreak risks** (e.g., secondary winter waves before vaccination was available)  

Accordingly, **early epidemic arrival** is expected to correlate with **higher prevalence, reproduction rates, and excess mortality** at the national level.  

To test this hypothesis, the initial approach suggested by @neuwirth2024 (which I will evaluate and refine in subsequent sections) followed five key steps:

1. **Construct a Global Mobility Network (GMN)** using worldwide air travel data from early 2020 (**OpenSky COVID-19 Flight Dataset and/or FlightRadar24 API**) to represent the situation at the outbreak’s onset.  
2. **Implement a SIR network simulation model** as presented by @brockmann2013.  
3. **Run outbreak simulations** (seeding in Wuhan) to estimate **arrival times (ranks)** of the epidemic in different regions.  
4. **Validate the model** by comparing simulated epidemic arrivals with reported data.
5. **Correlate modeled epidemic arrivals with reported excess mortality**, assessing the impact of arrival timing on overall mortality outcomes.

*However, initial evaluation of data sources reveals several practical and methodological challenges, potentially limiting the feasibility of a fully global analysis. I will address these issues and propose methodological refinements in subsequent sections. Finally, showcase [my own Hypothesis](#hypo)*

## Data Feasibility and Coverage

This section outlines my evaluation of publicly accessible aviation data sources, with the aim of assessing whether a global SIR model based on air travel is realistically implementable within the scope of a bachelor’s thesis. The findings justify a Europe-focused network as the most feasible and methodologically sound approach.

Although the initial approach proposed by @neuwirth2024 emphasizes a global-scale simulation using OpenSky or FlightRadar24, preliminary explorations and data evaluations have revealed substantial feasibility challenges. This section addresses key limitations encountered, particularly regarding **data accessibility, coverage issues**, and **practical modeling constraints**.

Specifically, crowdsourced ADS-B datasets such as OpenSky present **significant coverage gaps** in critical regions like China, especially during the early months of the pandemic (e.g., February 2020). Conversely, high-quality commercial datasets such as Eurocontrol, while robust, provide only **limited temporal coverage** (four discrete months per year), complicating continuous epidemic modeling.

### February 2020 Flight Data

During February 2020, the raw OpenSky dataset contained **2,648,835 flight records** (@fig-plot1, top panel). After discarding incomplete entries—flights lacking either an origin or a destination—**1,617,962 records** remained, reflecting a data reduction of **38.92%** (@fig-plot1, bottom panel). Among these cleaned entries, **674 flights** originated in **Mainland China, Hong Kong, or Macau** and terminated within the [Eurocontrol zone's member states](https://www.eurocontrol.int/our-member-and-comprehensive-agreement-states) (excluding Iceland, which joined in 2025). This modest number illustrates the **limited direct connectivity** captured by OpenSky during this critical month. While likely an **underestimate** of actual passenger flows due to regional coverage gaps, this subset nonetheless provides a valuable **baseline** for modeling early epidemic arrival into Europe.

```{r}
#| label: define_zones
#| echo: false
#| warning: false
library(tidyverse)
library(patchwork)

df <- read_csv("data/opensky/opensky_2020_02.csv.gz")
airports <- read_csv("data/airports.csv")

eurocontrol_countries <- c("BE", "FR", "DE", "LU", "NL", "GB", "IE", "PT", "GR", 
                            "MT", "TR", "CY", "HU", "CH", "AT", "DK", "NO", "SI", 
                            "SE", "CZ", "IT", "RO", "BG", "HR", "MC", "SK", "ES", 
                            "MK", "MD", "FI", "AL", "BA", "PL", "UA", "RS", "AM", 
                            "LT", "ME", "LV", "GE", "EE")

# Extract ICAO codes for China, HK, MO
china_airports <- airports %>%
  filter(iso_country %in% c("CN", "HK", "MO"), !is.na(icao_code)) %>%
  pull(icao_code)

europe_airports <- airports %>%
  filter(iso_country %in% eurocontrol_countries, !is.na(icao_code)) %>%
  pull(icao_code)

df_cleaned <- df %>%
  filter(!is.na(origin), !is.na(destination))

total_flights <- nrow(df)
cleaned_flights <- nrow(df_cleaned)
removed_pct <- round((1 - cleaned_flights / total_flights) * 100, 2)

# Summarize flights from China -> Europe
china_to_europe_flights <- df_cleaned %>%
  filter(origin %in% china_airports, destination %in% europe_airports) %>%
  nrow()

raw_counts <- df %>% count(day)
cleaned_counts <- df_cleaned %>% count(day)

```


```{r}
#| label: fig-result_feb_table
#| echo: false
#| warning: false

summary_data <- data.frame(
  Metric = c(
    "Total flights (raw)", 
    "Total flights (cleaned)", 
    "Percentage removed", 
    "China → Eurocontrol flights"
  ),
  Value = c(
    total_flights, 
    cleaned_flights, 
    paste0(removed_pct, "%"), 
    china_to_europe_flights
  )
)

knitr::kable(summary_data, caption = "Summary of flight data for February 2020.")
```

```{r}
#| label: fig-plot1
#| fig-cap: "Daily flight counts in February 2020 before (top) and after (bottom) data cleaning. The raw dataset (red line) included flights with missing origins or destinations, while the cleaned dataset (blue line) excludes incomplete entries."
#| echo: false
#| warning: false
p_raw <- ggplot(raw_counts, aes(x = day, y = n)) +
  geom_line(color = "red") +
  labs(title = "Daily Flight Counts (Raw Data)", x = "Date", y = "Flights")

p_cleaned <- ggplot(cleaned_counts, aes(x = day, y = n)) +
  geom_line(color = "blue") +
  labs(title = "Daily Flight Counts (Cleaned Data)", x = "Date", y = "Flights")

print(p_raw / p_cleaned)

```

To overcome these limitations, I propose integrating **multiple complementary data sources**. Specifically, the **OpenSky** dataset from February 2020—despite its limited volume—will serve as a **baseline** for the early epidemic arrival phase. **Eurocontrol** data (starting March 2020), offering comprehensive European flight records, will help validate February’s flight numbers and fill key gaps. For months lacking explicit records, **linear interpolation** and **seasonal normalization** methods may be employed, ensuring continuity in flight data for the initial seeding period. Recognizing early travel restrictions in China (e.g., Wuhan’s lockdown—see [here](#wavetime)) further highlights the importance of including multiple major Chinese airports, rather than focusing on Wuhan alone.

### Crowdsourced ADS-B: Strengths and Weaknesses

Multiple studies have leveraged crowdsourced ADS-B data for COVID-19 mobility analyses [@suzumura2020; @iacus2020; @salgas2024], yet coverage remains heavily biased toward North America and Europe. For instance, @suzumura2020 analyzed ~1.9 million flights (1 January–30 April 2020) from OpenSky’s dataset but noted minimal representation of flights into or out of China. By contrast, @strohmeier2021 reports that OpenSky recorded 30,989,481 flights in 2019—roughly 45% of all global flights that year—still undercounting major regions such as China, Africa, and parts of Asia. Consequently, reconstructing Wuhan-based travel for early 2020 remains particularly challenging:

> “OpenSky Network [...] does not provide global coverage and [...] flight data for major global airports such as the Beijing Capital International Airport in China and the Soekarno-Hatta International Airport in Indonesia are not available.” [@iacus2020]

Such large discrepancies between partial flight samples from @suzumura2020 (early 2020 subset) and near-full annual coverage by @strohmeier2021 (2019 totals) underscore the network’s highly uneven global sensor distribution. While some countries enjoy excellent receiver density, enabling robust analysis, large swaths of Asia and Africa have insufficient coverage. This unevenness must be accounted for in any study that relies on ADS-B data for measuring pandemic-related changes in global air travel. Alternatives mentioned below may help to mitigate these gaps, either by incorporating proprietary datasets (OAG/IATA), leveraging regional traffic records, or exploring hybrid machine-learning approaches using open data.

*Given these coverage biases, researchers often look beyond crowdsourced ADS-B to fill critical data gaps, as discussed next.*

### Proprietary Data and Regional Sources

Given the substantial coverage gaps identified with crowdsourced ADS-B sources, especially concerning early 2020 flights originating in China, researchers often turn to proprietary or regional data providers as more comprehensive alternatives. For instance, the **FlightRadar24 (FR24)** platform, highlighted by @neuwirth2024, offers extensive aviation data through a commercial ADS-B and multilateration network. @quadros2022 successfully utilized FR24 data to compile global civil aviation emissions between 2017 and 2020, demonstrating the potential for capturing diverse flight types—including cargo and unscheduled flights—that traditional schedule-only databases (such as OAG) might omit.

However, FR24 data are also subject to notable constraints:

- Coverage remains uneven, particularly for periods prior to 2020 and in regions such as Asia and Africa [@quadros2022].
- Data retrieval involves significant computational overhead, as large-scale queries for historical flight positions require substantial processing resources.
- Cost constraints pose a major barrier. FR24 employs a subscription-based, credit-driven pricing structure:
  - **Essential tier:** $90/month, offering 333,000 monthly credits and 2-year historical access.
  - **Advanced tier:** $900/month, providing 4,050,000 credits and unlimited historical access (subject to system limitations).

While comprehensive commercial databases (e.g., OAG or IATA) are available and offer near-complete coverage of scheduled flights and passenger flows, they are similarly subscription-based and costly, limiting accessibility primarily to well-funded research projects [@chinazzi2020]. As a result, many researchers adopt a regional approach, using datasets like **Eurocontrol** (Europe) or **FAA** (U.S.).

Considering these factors, a preliminary exploration of both Eurocontrol and OpenSky datasets was conducted. While OpenSky data from February 2020 included only **674 flights from China to Europe as previously stated.**, Eurocontrol’s data, although comprehensive for specific months (e.g., March, June, August, December), lacked crucial coverage for February 2020—the critical early period of COVID-19’s initial spread to Europe (first detected in France). To address these limitations, a hybrid data integration approach—combining Eurocontrol’s monthly snapshots with OpenSky’s February data—was assessed. This approach would necessitate interpolation of missing monthly flight counts and normalization across datasets, ensuring temporal continuity and data consistency for modeling.

Given the resource-intensity, uneven coverage, and potential analytical limitations posed by global proprietary data, this thesis will adopt a Europe-focused regional approach. By leveraging Eurocontrol’s detailed datasets, supplemented and temporally harmonized with OpenSky data for missing intervals (e.g., February 2020), the analysis will achieve sufficient spatial-temporal granularity to robustly model the early epidemic spread into Europe.

### ADS-B Exchange Archives

**ADS-B Exchange** offers **high-resolution (5-second interval)** global flight data from **March 2020 onward**. Although this represents an improvement over other open sources, critical coverage gaps persist, especially for **mainland China and specifically Wuhan**, significantly limiting its usefulness for capturing the pandemic's initial spread phase [@salgas2024]. Since the primary research question of this thesis concerns the early months of 2020 (January and February), ADS-B Exchange cannot adequately fill these crucial temporal gaps.

*Thus, despite its detailed nature, the post-March 2020 coverage of ADS-B Exchange is insufficient to study the earliest phases of COVID-19 transmission relevant to this research.*

### A Different Approach: Wikipedia + XGBoost

A promising open-data alternative to overcome gaps in aviation datasets has been recently proposed by @salgas2024. Their method merges **Wikipedia-scraped airport route data** with **XGBoost-based seat-capacity estimations**, providing a novel way to impute missing flight connections. Although originally applied to **2019 data**, this approach theoretically could be extended to early 2020 using historical Wikipedia snapshots (e.g., via the Internet Archive).

However, as emphasized by @salgas2024, this method requires substantial effort, including:

- manual Wikipedia data extraction,
- significant data cleaning,
- iterative training and validation of the machine learning model, and
- careful integration with radar-based or administrative flight records.

These processes are described as **"long and complex"** and yield results primarily suitable for **aggregated analyses at national or continental levels**, rather than precise route-level accuracy. Given the practical scope and limited available hours (~300 hours) for this bachelor's thesis, adopting this full-scale Wikipedia + XGBoost pipeline would exceed project constraints. Therefore, while innovative and valuable for future research endeavors requiring comprehensive global data, this approach is considered beyond the feasible scope of the current thesis.

*Thus, despite its methodological innovation, the Wikipedia + XGBoost strategy is judged impractical for this project's timeframe and resource constraints.*

### Seats vs. Actual Passengers {#ghost}

Another notable methodological challenge arises from the discrepancy between available data (typically seat capacities) and actual passenger numbers. Most **open-source flight datasets** (e.g., OpenSky, ADS-B Exchange) lack explicit passenger information, leading researchers to estimate traveler numbers by applying **assumed passenger load factors (PLF)** to seat capacities.

Pre-pandemic, the global PLF reached a record **82.6% in 2019** [@IATA2020], yet dropped sharply to **65.2% in 2020** due to COVID-19 [@Statista2024]. This rapid decline introduced additional uncertainty into passenger-flow estimates. The main inaccuracies include:

- **Cargo flights misidentified as passenger flights**: Unlabeled cargo-only operations may mistakenly inflate passenger estimates if treated as passenger flights [@suzumura2020].
- **Reduced occupancy and ghost flights**: Many flights operated near-empty to maintain airport slots or for regulatory purposes [@sun2022_b], complicating the reliability of simple PLF approximations.

*Given the data availability constraints, the common practice of multiplying seat capacities by historical or assumed PLFs remains necessary, despite acknowledging these inherent limitations. This thesis explicitly recognizes potential inaccuracies and will incorporate sensitivity analyses on these load-factor assumptions where feasible.*

### Are Passengers “Isolated” in Transit? How Does This Affect the Model?

A further simplification inherent to large-scale network models, such as the **SIR** model presented by @brockmann2013, involves treating air travel as instantaneous transmission events:

- **No in-flight quarantine or detailed cabin modeling**: Passengers are not considered isolated during flights; the model effectively assumes immediate infection potential upon arrival.
- **Ignored intra-cabin transmission**: While cabin-specific airflow and seating arrangements can influence in-flight transmission [@wang2023], these micro-level factors are typically omitted in macro-scale epidemic models due to computational and data complexity.

Thus, flights function primarily as **network links facilitating long-distance virus introductions**, rather than detailed sites of infection dynamics.

**In conclusion**, given these significant limitations with existing open (ADS-B Exchange, Wikipedia + XGBoost) and proprietary datasets (FR24, OAG/IATA), the initial global epidemic modeling proposal is pragmatically refined into a regional, Europe-centric approach. Leveraging robust Eurocontrol datasets supplemented by strategically selected OpenSky records, this thesis employs careful temporal interpolation, normalization, and sensitivity analyses. This adapted methodology aligns realistically with available resources, ensuring robust and feasible analysis of how early epidemic arrival influenced excess mortality across European nations.

**The next section** outlines the practical implementation steps of integrating this refined data strategy within a network-based SIR epidemic modeling framework.

## Rationale and Potential Approaches for Future Research

### Regional vs. Global Scope

Given the substantial limitations of available aviation data, a key methodological decision revolves around selecting an appropriate geographic scope. A **global approach**, while appealing due to its comprehensiveness, faces significant data challenges, particularly in regions with poor ADS-B coverage, such as China, Southeast Asia, and Africa [@suzumura2020; @iacus2020]. To mitigate these coverage gaps, one could employ methods such as Wikipedia scraping combined with XGBoost-based seat-capacity estimations to partially fill data voids [@salgas2024]. However, as previously highlighted, such an approach involves substantial labor overhead and complexity, making it less suitable for short-term projects.

In contrast, adopting a **regional scope**, particularly **Europe**, offers several distinct advantages:

- **Robust data availability** through the Eurocontrol network, which provides detailed flight records within Europe.
- Greater methodological reliability, as Europe’s aviation patterns are consistently captured by multiple data sources (e.g., Eurocontrol complemented by OpenSky).
- Practical alignment with project constraints (~300 hours available), allowing for thorough and well-founded analysis without sacrificing accuracy.

Moreover, intra-country transmission dynamics also play a significant role in epidemic spread. For instance, @laroze2021 showed that domestic commuting patterns significantly influenced COVID-19’s progression within English districts. A Europe-centric modeling approach thus presents an opportunity to incorporate not only international but potentially domestic and subregional mobility patterns as well, thereby capturing a more realistic spread scenario within manageable data requirements.

Given these considerations, **this thesis adopts a Europe-centric network approach**, enabling precise analysis of epidemic arrival timing and its subsequent impact on national-level excess mortality. The next sections will detail the methodological steps taken to integrate these European flight data into a network-based epidemic simulation.

### SIR as a Model Implementation

The thesis employs a **network-based SIR model**, as originally presented by @brockmann2013, to simulate the initial epidemic spread through aviation networks. Although ideally Wuhan airport would serve as the starting node (initial seeding), early data limitations make it necessary to consider other major Chinese airports (e.g., Shanghai, Beijing, Shenzhen) as proxies, especially given Wuhan's strict travel restrictions beginning 23 January 2020 [@chinazzi2020]. This approach aligns with previous studies highlighting the central role of major international hubs in early global dissemination [@lai2020].

Moreover, **Wuhan** itself is widely recognized as a major internal transport nexus [@zhang2020_a; @zhu2019], with air passenger flows strongly correlated to outbreak severity in connected regions [@zhao2020; @ai2020]. However, since multiple modes (rail, road, air) facilitated COVID-19’s domestic spread, focusing on Wuhan airport alone may underestimate case exportations—particularly after the 23 January lockdown. Incorporating data from other large airports better captures the broader seeding routes, ensuring our model reflects the reality that multiple Chinese hubs likely contributed to Europe’s early epidemic arrival.

A more sophisticated modeling choice would involve using a compartmental **SEIR** (or multi-stage **SEnIR**) approach, capturing a more realistic incubation period distribution (Gamma-like) and enhancing epidemiological accuracy [@abou2020; @rozan2024]. Such models provide nuanced insights into epidemic peaks, spread timing, and infection dynamics. However, they require additional parameters, detailed calibration, and more extensive computational resources, often demanding agent-based or fine-grained simulations.

*Given this thesis’s resource constraints, the simpler **SIR formulation**—using a mean-rate approximation of incubation and recovery periods as suggested by @brockmann2013 is methodologically sufficient to estimate initial arrival ranks of the epidemic across Europe. Nevertheless, future extensions of this work could explore implementing the more detailed **SEnIR** approach, particularly for studies aiming to examine secondary waves, detailed transmission dynamics, or public health interventions over longer time frames.*

## Specified Data

Many providers of flight movement data (e.g., FR24, OpenSky, ADS-B Exchange) rely heavily on **crowdsourced ADS-B** signals, leading to uneven coverage and potential biases in regions with fewer ADS-B receivers. In contrast, **Eurocontrol** gathers **flight plans and updates for all IFR** (Instrument Flight Rules) flights within European airspace, integrating radar data and other Air Navigation Service Providers (ANSPs) sources to offer a centralized, **operationally verified** overview.

### Access & Terms of Service

- Eurocontrol grants access via an **R&D Platform** exclusively for **non-commercial research**, requiring **user registration** and acceptance of their Terms of Service (TOS).  
- While **raw ATM data** sharing is prohibited, researchers can publicly disseminate **aggregated or derived** insights (e.g., charts, maps, diagrams) provided Eurocontrol is appropriately **credited**.  
- For this thesis, I obtained **personal access** under strict adherence to these research-focused guidelines.

### Airport Reference Data

To match flight records with actual airport locations and metadata (including ICAO codes, IATA codes, lat/long coordinates, and airport categories), this thesis uses the publicly available CSV data provided by [**OurAirports**](https://ourairports.com/). Although not formally required by OurAirports’ terms of use, appropriate credit is given here in recognition of their comprehensive dataset, which is updated daily and includes worldwide airports, heliports, and other facilities. The file used in this project is `airports.csv` (dated 24 March 2025), obtained via the **OurAirports** [Website](https://ourairports.com/data/).

### Data Availability & Format

- Eurocontrol’s dataset covers **four snapshot months per year**: March, June, September, and December.
- It includes **commercial flights only** (types `S` for scheduled and `N` for non-scheduled), excluding general aviation and military operations.
- Detailed flight-level information comprises scheduled and actual departure/arrival times, aircraft type (ICAO 8643), operator details (ICAO 8585), route waypoints, and airspace specifics, updated each AIRAC cycle.
- Given the limited temporal coverage (four months annually), supplementary sources (e.g., **OpenSky**) will help fill critical temporal gaps.

> **Metadata & TOS:** Further details available via the [EUROCONTROL ATM R&D Datasets – Terms of Use](https://www.eurocontrol.int/publication/eurocontrol-atm-rd-datasets-terms-use). All data handling complies with **non-commercial research use**, restricting public dissemination of raw flight records.

### Combining with OpenSky

To bridge temporal gaps between the quarterly Eurocontrol snapshots, the thesis leverages the **OpenSky Network** dataset ("Crowdsourced air traffic data from The OpenSky Network 2020," @olive2023). Despite known coverage limitations, particularly in regions like China or Africa, OpenSky’s **continuous monthly data** provides essential continuity, enabling coherent modeling of air traffic trends preceding and following Eurocontrol snapshots.

### Modeling Europe’s Initial COVID-19 Wave and Time Windows {#wavetime}

Although **officially recognized** European COVID-19 cases did not appear until January 2020, at least one retrospective study suggests that SARS-CoV-2 may have reached Europe as early as **late December 2019** [@deslandes2020]. In that investigation, a single archived respiratory sample from a patient in the Paris region tested positive for SARS-CoV-2 upon re-analysis, hinting at possible sporadic introductions prior to January. However, the authors acknowledge that this result does not confirm large-scale circulation at the time.

By most accounts, the **major viral seeding** in Europe occurred through air travel from China throughout **January and February 2020** [@chinazzi2020]. Gradual travel restrictions emerged over this period:

- **23 January 2020:** Wuhan city enforced an internal lockdown, yet international travel continued, allowing asymptomatic or undetected carriers to reach global aviation hubs. As highlighted by @chinazzi2020, the travel quarantine of Wuhan delayed epidemic progression by only 3 to 5 days within mainland China but significantly reduced international case importations by nearly 80% until mid-February.  
  - Further modeling by @chinazzi2020 suggests that even **drastic travel reductions** (up to 90%) alone could delay—but not fully prevent—virus exportation unless paired with substantial reductions in local transmissibility (e.g., effective NPIs).

- **Late January–Early February 2020:** Several countries introduced formal travel bans on direct flights from China—for instance, the U.S. (31 January) and Australia (1 February)—although indirect transit through international hubs (e.g., Dubai, Doha) remained open [@whitehouse2020proclamation; @australia2020ban; @sun2020]. This partial restriction meant travelers could still route through third countries, underscoring the ongoing importance of global transit hubs during February.

- **March 2020:** Comprehensive lockdown measures across Europe dramatically curtailed flight volumes, but significant viral entry had already occurred by this stage, seeding community transmission.

- **8 April 2020:** After approximately 10 weeks of strict lockdown measures, Wuhan resumed air and train services, further underscoring how interconnected aviation networks can accelerate or revive international spread [@zhang2020_a].

Because **Eurocontrol** only provides quarterly snapshots, this thesis strategically integrates:

- **OpenSky (January–February, April 2020):** Documents overall flight trends and mobility changes prior to and following initial lockdowns.

- **Eurocontrol (March 2020):** Provides detailed flight data during Europe’s most intense lockdown phase, allowing for a more granular look at flight patterns in the critical month when widespread restrictions took hold.

By combining these sources, the analysis can partially account for the heightened transmission risk in **early 2020**, even if flight data coverage remains incomplete for certain Asian regions.

*While global transit hubs outside Europe (e.g., Dubai, Doha) remained critical in February for indirect arrivals, major European airports (e.g., Frankfurt, Paris Charles de Gaulle, London Heathrow) also served as high-volume nodes for intra-European distribution. Accounting for every potential route and transit point beyond April would vastly exceed the scope of this thesis. Instead, this study focuses on initial external seeding pathways from China—alongside aggregated intra-European flows—using OpenSky (January–February, April) and Eurocontrol (March) data, capturing the crucial early wave of viral importation and the subsequent lockdown effects without venturing into later pandemic phases.*

### Correlation with Excess Mortality

After simulating epidemic arrival timing in European countries using the adjusted SIR model, these predictions will be correlated specifically with national-level excess mortality rates to quantify whether early epidemic seeding systematically influenced mortality outcomes. [@theeconomist2021]

A recent short communication by Neuwirth & Elixhauser (202x, in preparation) compiles global excess mortality data but explicitly highlights that the most comprehensive and consistently measured datasets—including excess mortality alongside key socioeconomic indicators (e.g., GDP, healthcare quality, vaccination rates)—are predominantly available for European countries. Thus, despite the broader global intent of our dataset, the density and reliability of European data specifically justify the focused analysis undertaken here.

*Although incorporating additional socioeconomic factors would enrich the analysis, such an approach exceeds the scope and time constraints of this thesis. Nevertheless, the proposed Europe-focused correlation offers meaningful insights into how epidemic arrival timing may have impacted national mortality outcomes, contributing to a deeper understanding of early COVID-19 pandemic dynamics.*

### Excess Mortality Data

To measure country-level excess mortality during the first wave, this thesis uses estimates published by The Economist [@theeconomist2021], who constructed a machine-learning model to impute missing data and approximate total excess deaths, as described in [their explanatory article](https://www.economist.com/graphic-detail/2021/05/13/how-we-estimated-the-true-death-toll-of-the-pandemic). 

Although these estimates carry inherent uncertainties—particularly for countries with underreported statistics—this dataset offers a reasonably consistent, cross-national view of pandemic-attributable mortality. The analysis in this thesis uses the aggregated excess-death estimates through [January - April], focusing on the first COVID-19 wave in Europe.

## Advanced Research Question & Hypothesis {#hypo}

**Research Question**

> Does the timing of epidemic seeding—derived from integrated air traffic data (OpenSky for January–February and April–May, and Eurocontrol for March and June 2020)—correlate with higher excess mortality in Europe during the first wave of COVID-19?

**Hypothesis**  
We hypothesize that **countries experiencing earlier epidemic seeding**, as inferred from our integrated flight network analysis (January–April 2020), will exhibit **higher first-wave excess mortality**. Although controlling for additional socioeconomic factors would strengthen the analysis, this thesis explicitly focuses on the direct relationship between epidemic arrival timing and mortality, acknowledging socioeconomic elements as important but beyond the immediate scope.

The anticipated mechanism driving this relationship includes:

- **Shorter reaction times for interventions:**  
  Countries seeded earlier had less opportunity to implement timely non-pharmaceutical interventions (NPIs), potentially leading to more rapid spread and higher mortality.

- **Increased strain on healthcare systems:**  
  Early epidemic waves could overwhelm healthcare resources, exacerbating mortality rates.

- **Heightened exposure through international hubs:**  
  Major airports (e.g., Frankfurt, Paris, London) potentially acted as primary gateways for viral entry, rapidly amplifying epidemic spread into densely populated regions.

By integrating OpenSky and Eurocontrol data, our Europe-centric model will effectively capture the key patterns and timing of epidemic seeding events, allowing for a meaningful analysis of their potential correlation with excess mortality.


### Research Design and Methodology

In line with my **Supervisor’s advice (March 14, 2025)**, this thesis **avoids complex agent-based modeling (ABM)** in favor of a **simplified network-based SIR model** [@brockmann2013]. This streamlined approach is justified by:

- **Feasibility:** Avoids excessive complexity, aligning with the limited scope and timeline of a bachelor’s thesis.  
- **Aggregated Flows:** Airports (and ultimately countries) are represented as **nodes**, with aggregated passenger flows as edges—maintaining manageable data complexity.  
- **Preservation of Core Dynamics:** Captures critical aspects needed to investigate epidemic seeding patterns and mortality relationships *without* intricate individual-level modeling.

### Specific Methodological Steps

1. **Construct a European Mobility Network (EMN)**  
   - **Prioritize European airports** based on data coverage (Eurocontrol/OpenSky), emphasizing major hubs (e.g., Paris, London, Frankfurt) that capture significant passenger volumes.  
   - **Approximate passenger flows** between airports using *seat capacity × an assumed load factor*.  
     - An **important Note:** This approach may overestimate passenger totals if [ghost flights or cargo-only flights are present](#ghost), or if actual load factors differ regionally. A brief sensitivity check (e.g., 50%, 70%, 90% load factor) will therefore demonstrate the robustness of results.
   - **Visualization:** Create simplified yet illustrative airport-level visualizations (e.g., Sankey diagrams). For statistical analyses, flows are **aggregated at the country level**.

2. **Implement a Simplified ODE-Based SIR Model**  
   - Represent each airport as a node with daily changes in **Susceptible (S), Infected (I), Recovered (R)** compartments.  
   - Account for **all relevant daily flights** (both from China and **within** Europe) by aggregating seat capacity × assumed load factor, thereby simulating how infections spread among European nodes once seeded.  
   - Rather than using Wuhan alone as the origin, incorporate inbound flight data from **multiple major Chinese airports** (e.g., Beijing, Shanghai, Shenzhen) to approximate early European seeding, consistent with research showing widespread exportation after Wuhan’s 23 January 2020 lockdown [@chinazzi2020; @zhang2020_a].

3. **Estimate Country-Level Epidemic Arrival Times**  
   - Define an **infection threshold** for “arrival time” at each airport.  
   - Aggregate airport-level arrivals to the **country level**, yielding an epidemic arrival timeline per country.

4. **Rank Countries by Estimated Epidemic Arrival**  
   - Sort countries from earliest to latest based on aggregated arrival times, indicating each country’s relative position in the seeding timeline.

5. **Correlation with Excess Mortality**  
   - **Spearman’s rank correlation**: Compare each country’s arrival rank with its **excess mortality** (during the first-wave period).  
   - Emphasize that **no socioeconomic controls** (e.g., GDP, healthcare capacity) are included in this initial approach, so the correlation should be considered exploratory.

### Project Timeline

## References

::: {#refs}
:::
