---
title:  "Inbound flights × excess mortality — correlation results"
author: "M. Elixhauser"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
freeze: auto
execute:
  echo: false 
---

```{r setup, include = FALSE}
source(here::here("R", "00_load_libs.R"))

fig_dir <- here("data/figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

analysis_df <- read_rds(here("data/processed/analysis_df.rds"))

pop_df <- read_csv(
  here("data/raw/owid/owid-covid-data.csv"),
  show_col_types = FALSE,
  col_select     = c(iso_code, date, population)
) |>
  filter(lubridate::year(date) == 2020) |>
  group_by(iso_code) |>
  slice_max(date, with_ties = FALSE) |>
  ungroup() |>
  transmute(iso3 = iso_code, population)
```

## Analysis data set

```{r}
corr_df <- analysis_df |>
  semi_join(analysis_df |> count(iso3) |> filter(n == 4), by = "iso3") |>
  left_join(pop_df, by = "iso3") |>
  mutate(flights_pm = total_inbound_flights_combined / (population / 1e6))
stopifnot(!anyNA(corr_df$flights_pm))
```

## Spearman's rank correlation (baseline)

```{r}
boot_rho <- function(data, idx, xvar) {
  with(
    data[idx, ],
    cor(get(xvar),
      excess_mortality_cumulative_per_million,
      method = "spearman",
      use = "complete.obs"
    )
  )
}

spearman_by_year <- function(xvar, B = 5000) { # Bootstrap = 5000 - Default .. need a reasoning though.

  corr_df %>%
    group_by(target_date) %>%
    group_modify(\(d, key) {
      ok <- complete.cases(
        d[[xvar]],
        d$excess_mortality_cumulative_per_million
      )
      dd <- d[ok, ]
      n <- nrow(dd)

      if (n < 3 ||
        sd(dd[[xvar]]) == 0 ||
        sd(dd$excess_mortality_cumulative_per_million) == 0) {
        return(tibble(
          rho = NA, ci_lo = NA, ci_hi = NA,
          p = NA, n = n
        ))
      }

      rho <- cor(dd[[xvar]], dd$excess_mortality_cumulative_per_million,
        method = "spearman"
      )
      pval <- cor.test(dd[[xvar]], dd$excess_mortality_cumulative_per_million,
        method = "spearman", exact = FALSE
      )$p.value

      set.seed(42)
      ci <- if (n >= 4) {
        boot(dd, boot_rho, R = B, xvar = xvar) |>
          (\(b) quantile(b$t, c(.025, .975), na.rm = TRUE))() # .025, .975 CI (why used!)
      } else {
        c(NA, NA)
      }

      tibble(
        rho = rho,
        ci_lo = ci[1],
        ci_hi = ci[2],
        p = pval,
        n = n
      )
    }) %>%
    ungroup() %>%
    mutate(
      variable = xvar,
      year = lubridate::year(target_date)
    )
}

spearman_res <- purrr::map_dfr(
  c(
    "total_inbound_flights_dec19",
    "total_inbound_flights_mar20",
    "total_inbound_flights_combined",
    "flights_pm"
  ),
  spearman_by_year
) %>%
  mutate(variable = dplyr::recode(variable,
    total_inbound_flights_dec19    = "Dec 2019",
    total_inbound_flights_mar20    = "Mar 2020",
    total_inbound_flights_combined = "Combined",
    flights_pm                     = "Flights / M pop"
  ))
```


```{r}
knitr::kable(
  spearman_res |>
    select(year, variable, rho, ci_lo, ci_hi, p, n) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))),
  caption = "Table 4-1 Spearman ρ between early flight exposure and cumulative excess mortality."
)
```

**Key patterns:**
- Early air conenctivity explains a sizeable share of the cross-country variation in first-wave excess mortality. (p ≈ -0.45, 95 % CI 0.18-0.80; p < 0.01).
- From 2021 onwards the simple rank correlation vanishes; by 2022-23 it turns modestly negative (p ≈ -0.45, 95 % CI -0.79-0.00).
- The sign change is larger than the bootstrap confidence intervals, indicating that it is unlikely to be due to sampling noise in a 25-country sample.

**Interpretation:**
International flights mattered only during the narrow "pre-transmission window" (need better term). Once SARS-CoV-2 was seeded "everywhere", subsequent excess mrotality was driven by domestic factors (age structure, public-health response, vaccination, health-system strain) rather than by additional import pressure. This finding is ... (search for SOURCE in: diminishing return or not of border closures after local circulation is established.. could be interesting)

# Visualizing the trend

## p over time
```{r}
p_rho <- ggplot(
  spearman_res,
  aes(target_date, rho, colour = variable, group = variable)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_x_date(date_labels = "%Y") +
  labs(x = NULL, y = "Spearman ρ", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_rho |> (\(p) {
  ggsave(
    filename = here(fig_dir, "rho_timeseries.png"),
    plot = p,
    width = 6, height = 4, dpi = 300, bg = "transparent"
  )
  p
})()
```

**Take-away:**

- Import effect is early and short-lived. Direct flights explain ≈ 3 % of the rank variance in May 2020, but < 10 % (< 5 % once micro‑states are excluded) from 2021 onwards. …
- Once the virus is seeded, domestic factors dominate. The mere fact of being well-connected n longer increases cumulative mrotality and may evne correlate with lower late-period deaths (?)
- Policy implciation (exploratory) Border controls can delay the first wave but have little leverage on the long-run toll unelss coupled with sustained internal measures.

## Snapshot scatter (05 May 2020)
```{r}
p20 <- corr_df %>%
  filter(target_date == as.Date("2020-05-05")) %>%
  ggplot(aes(
    total_inbound_flights_combined,
    excess_mortality_cumulative_per_million
  )) +
  geom_point(size = 3, alpha = .8, colour = "#2b8cbe") +
  geom_text_repel(aes(label = iso3), max.overlaps = 12, size = 3) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Inbound flights (log-scale)",
    y = "Excess deaths / million",
    caption = "ρ = 0.53 (p ≈ 0.007)"
  )

p20
ggsave(here(fig_dir, "scatter_2020.png"),
  p20,
  width = 5.6, height = 4, dpi = 300
)
```

## pp scatter (Inb flights, dec 2019 + mar 2020) - with highlighter 
- planned! 

## Scater by year

# Robustness check

# Takeaway
