---
title:  "Inbound flights × excess mortality — correlation results"
author: "M. Elixhauser"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
freeze: auto
execute:
  echo: false 
---

```{r setup, include = FALSE}
source(here::here("R", "00_load_libs.R"))

fig_dir <- here("data/figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

analysis_df <- read_rds(here("data/processed/analysis_df.rds"))

pop_df <- read_csv(
  here("data/raw/owid/owid-covid-data.csv"),
  show_col_types = FALSE,
  col_select     = c(iso_code, date, population)
) |>
  filter(lubridate::year(date) == 2020) |>
  group_by(iso_code) |>
  slice_max(date, with_ties = FALSE) |>
  ungroup() |>
  transmute(iso3 = iso_code, population)
```

## Analysis data set

```{r}
corr_df <- analysis_df |>
  semi_join(analysis_df |> count(iso3) |> filter(n == 4), by = "iso3") |> # Keep only countries with complete data (all 4 analysis dates)
  left_join(pop_df, by = "iso3") |>
  mutate(flights_pm = total_inbound_flights_combined / (population / 1e6))
stopifnot(!anyNA(corr_df$flights_pm))
```

## Spearman's rank correlation (baseline)

```{r}
# Bootstrapped Spearman rho for robust CIs (idx is bootstrap sample indices)
boot_rho <- function(data, idx, xvar) {
  with(
    data[idx, ],
    cor(get(xvar),
      excess_mortality_cumulative_per_million,
      method = "spearman",
      use = "complete.obs"
    )
  )
}

spearman_by_year <- function(xvar, B = 5000) {
  corr_df |> 
    group_by(target_date) |> 
    group_modify(\(d, key) {
      ok <- complete.cases(
        d[[xvar]],
        d$excess_mortality_cumulative_per_million
      )
      dd <- d[ok, ]
      n <- nrow(dd)

      if (n < 3 ||
        sd(dd[[xvar]]) == 0 ||
        sd(dd$excess_mortality_cumulative_per_million) == 0) {
        return(tibble(
          rho = NA, ci_lo = NA, ci_hi = NA,
          p = NA, n = n
        ))
      }

      rho <- cor(dd[[xvar]], dd$excess_mortality_cumulative_per_million,
        method = "spearman"
      )
      pval <- cor.test(dd[[xvar]], dd$excess_mortality_cumulative_per_million,
        method = "spearman", exact = FALSE
      )$p.value

      set.seed(42)
      ci <- if (n >= 4) {
        boot(dd, boot_rho, R = B, xvar = xvar) |>
          (\(b) quantile(b$t, c(.025, .975), na.rm = TRUE))() # .025, .975 CI (why used!)
      } else {
        c(NA, NA)
      }

      tibble(
        rho = rho,
        ci_lo = ci[1],
        ci_hi = ci[2],
        p = pval,
        n = n
      )
    }) |> 
    ungroup() |> 
    mutate(
      variable = xvar,
      year = lubridate::year(target_date)
    )
}

spearman_res <- purrr::map_dfr(
  c(
    "total_inbound_flights_dec19",
    "total_inbound_flights_mar20",
    "total_inbound_flights_combined",
    "flights_pm"
  ),
  spearman_by_year
) %>%
  mutate(variable = dplyr::recode(variable,
    total_inbound_flights_dec19    = "Dec 2019",
    total_inbound_flights_mar20    = "Mar 2020",
    total_inbound_flights_combined = "Combined",
    flights_pm                     = "Flights / M pop"
  ))
```


```{r}
knitr::kable(
  spearman_res |>
    select(year, variable, rho, ci_lo, ci_hi, p, n) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))),
  caption = "Table X Spearman ρ between early flight exposure and cumulative excess mortality."
)
```

**Key patterns:**

* Early air connectivity explained a substantial share of cross-country variation in first-wave excess mortality (ρ ≈ 0.53, 95% CI \[0.18, 0.80], p < 0.01).
* From 2021 onwards, the rank correlation essentially disappeared; by 2022–23 it became modestly negative (ρ ≈ –0.45, 95% CI \[–0.79, 0.00]).
* This sign change exceeds the bootstrap confidence intervals, suggesting the shift is unlikely due to random sampling in a \~25-country dataset.
* Notably, **“Flights per million population” was already significantly negative from 2021 onward**, indicating that higher per-capita connectivity was linked to *lower* late-pandemic mortality—a possible “resource advantage” effect in highly connected countries.

**Interpretation:**

* International flights played a major role only during the initial seeding phase—before widespread community transmission in Europe.
* Once SARS-CoV-2 was established in most countries, *domestic factors*—population age, public health response, healthcare capacity, and vaccination coverage—became dominant in shaping excess mortality.
* This pattern aligns with studies showing that the effect of border controls diminishes once local transmission is established (see e.g., Chinazzi et al. 2020, *Science*).

**Why bootstrap CIs?**

* We computed 5,000 bootstrap resamples, following best practice for small-n correlation studies (DiCiccio & Efron 1996; Tong et al. 2016; Mokhtar 2023). This provides stable confidence intervals, and a fixed random seed (`set.seed(42)`) ensures full reproducibility.

**Summary:**
Early in the pandemic, countries most exposed to direct flights from China and Hong Kong suffered higher excess mortality. However, as the pandemic progressed, this association disappeared—and by 2022–2023, even reversed: high connectivity was linked to *lower* excess deaths, likely reflecting differences in domestic response, health system capacity, and perhaps a “rich country” advantage.
These findings underscore the narrow window during which travel restrictions can affect epidemic outcomes, and the overwhelming importance of internal factors once local transmission is widespread.


# Visualizing the trend

## p over time
```{r time_series, fig.cap="Time trend in the correlation (Spearman ρ) between early direct scheduled passenger flight exposure and excess mortality (2020–2023), for different flight metrics. Correlation is positive and statistically significant in 2020 (prevalent seeding effect), but vanishes or reverses in later years as domestic factors dominate pandemic outcomes."}
pal <- c(
  "Dec 2019" = "#E69F00",
  "Mar 2020" = "#56B4E9",
  "Combined" = "#009E73",
  "Flights / M pop" = "#CC79A7"
)

p_rho <- ggplot(
  spearman_res,
  aes(target_date, rho, colour = variable, group = variable)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  scale_colour_manual(values = pal, name = NULL) +
  scale_x_date(date_labels = "%Y") +
  labs(
    x = NULL,
    y = "Spearman ρ",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_rho |> (\(p) {
  ggsave(
    filename = here(fig_dir, "rho_timeseries.png"),
    plot = p,
    width = 6, height = 4, dpi = 300, bg = "transparent"
  )
  p
})()

```

Take-away:

* **Early impact, short window:** Direct flight connectivity strongly correlates with excess mortality during the first pandemic wave, suggesting early import pressure was a key driver of cross-country variation.

* **Rapid attenuation:** This association disappears after 2020 and reverses by 2022–2023, reflecting a shift from international importation to domestic drivers such as demographics, public health response, and health system capacity.

* **Policy implication:** Border controls can slow initial seeding but have limited effect once local transmission is established. Long-term pandemic outcomes depend more on internal mitigation and health system resilience.

## Snapshot scatter (05 May 2020)
```{r scatter_2020, fig.cap="Relationship between cumulative direct scheduled passenger flights from China and Hong Kong to each European country (x-axis, log-scale) and excess mortality per million residents by 5 May 2020 (y-axis). Each point is a country (ISO3 code). A moderate positive correlation is observed (Spearman ρ = 0.53, p ≈ 0.007), indicating higher early air connectivity was associated with greater first-wave excess mortality."}
p20 <- corr_df |> 
  filter(target_date == as.Date("2020-05-05")) |> 
  ggplot(aes(
    total_inbound_flights_combined,
    excess_mortality_cumulative_per_million
  )) +
  geom_point(size = 3, alpha = .8, colour = "#009E73") +
  geom_text_repel(aes(label = iso3), max.overlaps = 12, size = 3) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Direct scheduled passenger flights (log-scale)",
    y = "Excess deaths / million",
  )

p20
ggsave(here(fig_dir, "scatter_2020.png"),
  p20,
  width = 5.6, height = 4, dpi = 300
)
```

**Interpretation:**

Countries with the highest direct air connectivity to China and Hong Kong (such as Germany, the UK, France, and the Netherlands) experienced some of the highest excess mortality rates during the first wave of COVID-19. This positive rank correlation (ρ = 0.53) supports the hypothesis that early exposure via direct flights contributed significantly to cross-country differences in first-wave mortality. Outliers (e.g., Sweden, Ireland, Slovakia) may reflect unique domestic factors.

## pp scatter (Inb flights, dec 2019 + mar 2020)

```{r scatter_dec_mar, fig.cap = "Relationship between direct scheduled passenger flights from \ China and Hong Kong (log-scale, x-axis) and cumulative excess mortality per \ million residents measured on 5 May 2020 (y-axis). Panels separate the \ exposure metric for **December 2019** (left) and **March 2020** (right). A \ moderate positive association is visible in both months (Spearman ρ ≈ 0.50 \ and ρ ≈ 0.57). Country codes label (i) the three largest early-exposure hubs \ by flight count and (ii) countries in the top quintile of first-wave excess \ mortality, highlighting the states that drive the correlation." }

snapshots <- corr_df |>
  filter(target_date == as.Date("2020-05-05")) |>
  select(
    iso3, excess_mortality_cumulative_per_million,
    flights_dec19 = total_inbound_flights_dec19,
    flights_mar20 = total_inbound_flights_mar20
  ) |>
  pivot_longer(
    starts_with("flights_"),
    names_to  = "month",
    values_to = "inbound_flights"
  ) |>
  mutate(
    month = factor(
      recode(month,
             flights_dec19 = "Dec 2019",
             flights_mar20 = "Mar 2020"),
      levels = c("Dec 2019", "Mar 2020"))
  ) |>
  filter(inbound_flights > 0) # drop zero-flight countries

stats_lbl <- snapshots |>
  group_by(month) |>
  summarise(
    rho = cor(inbound_flights,
              excess_mortality_cumulative_per_million,
              method = "spearman")
  ) |>
  mutate(label = sprintf("ρ ≈ %.2f", rho),
         x = max(snapshots$inbound_flights)*0.9,
         y = min(snapshots$excess_mortality_cumulative_per_million)*0.9)

gg_pp <- ggplot(
  snapshots,
  aes(inbound_flights, excess_mortality_cumulative_per_million)
) +
  geom_point(size = 3, alpha = 0.8, colour = "#009E73") +
  geom_text_repel(
    data  = \(d) d |> filter(iso3 %in% c("ESP","ITA","GBR","DEU","BEL","FRA", "NLD")),
    aes(label = iso3),
    size  = 3
  ) +
  geom_text(
    data = stats_lbl,
    aes(x = x, y = y, label = label),
    hjust = 1, vjust = 0, size = 3, fontface = "bold",
    colour = "grey30"
  ) +
  scale_x_log10(
    breaks = c(1,3,10,30,100,300,1000),
    labels = scales::comma,
    expand = expansion(mult = c(.05, .02))
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = .03)
  ) +
  facet_wrap(~month, nrow = 1) +
  labs(
    x = "Direct scheduled passenger flights (log-scale)",
    y = "Excess deaths / million (to 5 May 2020)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text    = element_text(face = "bold"),
    panel.spacing = grid::unit(1, "cm")
  )

gg_pp

ggsave(
  here(fig_dir, "scatter_dec_mar.png"),
  gg_pp, width = 7.3, height = 3.8, dpi = 300, bg = "transparent"
)
```

**Interpretation:**

* **Consistent moderate association:** In both December 2019 and March 2020, countries with more direct scheduled passenger flights from China and Hong Kong tended to have higher excess mortality by early May 2020. This relationship is reflected in moderate Spearman correlations (ρ ≈ 0.50–0.57).

* **Pattern driven by hubs and hardest-hit countries:** The association is strongest for major Western European air hubs—such as Spain, Italy, the UK, Belgium, France, and the Netherlands—which were among the first and worst affected in the pandemic’s initial wave.

* **Interpretation:** These results support the hypothesis that early international connectivity played a significant role in seeding the first wave of SARS-CoV-2 across Europe. However, not all highly connected countries suffered high mortality, and not all countries with high mortality were major hubs—indicating that factors beyond air traffic (e.g., timing of interventions, population age, health system readiness) also shaped outcomes.

* **Caveat:** The observed association diminishes in later periods as local transmission and domestic factors become the dominant determinants of excess mortality, which aligns with global findings that border controls have the most impact only before widespread community spread is established.

## Caterpillar plot 

```{r caterpillar, fig.cap="Spearman ρ (±95% bootstrap CI) between early flight exposure and excess mortality, by reference year. The association is strongly positive in 2020, fades in 2021, and turns negative from 2022 onward."}

pal <- c(
  "Dec 2019" = "#E69F00",
  "Mar 2020" = "#56B4E9",
  "Combined" = "#009E73"
)

gg_cater <- spearman_res |>
  filter(variable %in% c("Dec 2019", "Mar 2020", "Combined")) |>
  mutate(variable = factor(variable,
                           levels = c("Combined", "Dec 2019", "Mar 2020"))) |>
  ggplot(aes(year, rho, colour = variable)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  geom_errorbar(
    aes(ymin = ci_lo, ymax = ci_hi),
    width = .15, linewidth = .9,
    position = position_dodge(width = .4)
  ) +
  geom_point(size = 3, position = position_dodge(width = .4)) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(x = NULL, y = "Spearman ρ (± 95 % CI)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

gg_cater

ggsave(here(fig_dir, "rho_caterpillar.png"),
       gg_cater, width = 6, height = 4, dpi = 300, bg = "transparent")

```

**Interpretation:**
* All three exposure metrics are clearly positive in 2020; in 2021 their CIs straddle zero, and from 2022 the combined metric is significantly negative, confirming a genuine sign-flip rather than sampling noise.

# Limitations {#limitations}

Can be used for Thesis.qmd

## Data limitations

**Flight-exposure proxy (EUROCONTROL, IFR plans).**
  Our exposure estimates count ICAO flight types S (scheduled) and N (non-scheduled commercial) as reported in EUROCONTROL's IFR data.
  -  Actual passenger numbers per flight are unknown; all filed flights—including "ghost" (very low-occupancy) and cargo-only flights—are counted equally.
  - Itineraries with connecting segments via third-country hubs (e.g., DXB, DOH) are not fully captured, so exposure is under-estimated for transfer states and may be over-estimated for major cargo airports such as Liège.
  - In 2020, Liège Airport handled over 1.1 million metric tons of airfreight (ranking 6th in Europe), with cargo volumes rising 23.5% year-on-year despite the pandemic—contrasting with heavy declines in passenger traffic at most airports [@fraport2021]. This highlights the distinct behavior of cargo airports, where flight counts and exposure indices remained elevated due to surging demand for freight and widespread use of "preighter" operations (passenger planes repurposed for cargo).
  - Only four "representative" months (Mar, Jun, Sep, Dec) are available per year from EUROCONTROL's data releases [@eurocontrolData2025], potentially missing key seeding windows (February 2020).

This limitation is directly reflected in our results: for example, Belgium’s exposure index does not drop in line with other countries during the pandemic period (**see Figure @pct_change_filtered**), consistent with Liège’s unique role as a freight hub.


**Ghost/Abnormal flights (load factor bias).**
  The phenomenon of "ghost flights"-nearly empty planes operated during the pandemic, often for slot-retention reasons-is well documented in Europe [@sun2022_b]. @sun2022_b show that such flights peaked in March-April 2020, especially among low-cost carriers and on frequently served intra-European routes.
  - Our exposure metric does not capture actual passenger loads or distinguish between ghost, cargo, or regularly filled flights, so "exposure" may be overstated for affected periods/routes.
  - @sun2022_b  focus on intra-European markets; the extent of "ghost flights" on China–Europe routes remains unclear, but similar slot-related pressures likely played a role early in the pandemic.

[!!!!! Research more about Chinese airlines and ghost flights, especially in the first months of 2020.]

[TODO: Read Warnock-Smith et al 2021 ]

* **Passenger vs. Cargo Dynamics: Exposure Index Bias at Cargo Hubs.**
  A core limitation of our flight-exposure proxy is that it cannot distinguish between passenger, cargo, and “ghost” flights (near-empty for slot retention), especially at major hubs. While most European airports saw catastrophic declines in passenger traffic during 2020 (e.g., Frankfurt -74%, [@fraport2021]), cargo hubs like Liège (+23.5% cargo volume year-over-year (YoY)) or Frankfurt (-8.3% cargo) showed much greater resilience—partly due to a surge in dedicated freighters and “preighter” (cargo-only) flights. At Frankfurt, over 8,600 preighter flights accounted for 8% of annual cargo, and freighters made up 80% of cargo movements.
  Consequently, flight counts and exposure indices may significantly overstate real human exposure at major cargo airports. This explains why indices for Belgium and Germany remained high despite catastrophic drops in passenger travel (see Figure @pct_change_filtered)—a pattern also seen at other major cargo hubs. Such discrepancies highlight the importance of careful interpretation when relying on flight-based proxies for pandemic exposure.

[HIGHLY IMPORTANT HERE: DESCRIPTIVE PLTOS THAT SHOWCASE DECLINE! BIG ONE! NOW I CAN EXPLAIN WHY GERMANY AND BELGIUM KEPT CONSISTENT!]
[See fraport and delcine of descirptive plots!]

Important Sections of @warnocksmith2021 for the Thesis:

> **NOTE:** The following content is **not original** and includes **verbatim excerpts** or close paraphrasing from @warnocksmith2021!!!

Air travel enables the rapidity of communicable disease transfer across international borders (Fadel et al., 2008) though cross-border travel by other modes (e.g. rail, road) can also lead to the rapid imported transmission of cases.

Neatly tied down: 
Li (2020) suggests that COVID-19 has a different influence on the Chinese market. Zhang et al. (2020a) investigated the risk of importing COVID-19 cases by foreign countries on Chinese provinces; they have not evaluated the traffic related implications to airlines and airports in China. Sun et al. (2020) provided a timely analysis of the network structures post COVID-19. Their spatial-temporal evolutionary dynamics approach looked at China, Europe and United States. Li (2020) focused on the cargo market and Czerny et al. (2020) examined the Chinese government's aviation policy choices in the light of the coronavirus pandemic. Meanwhile Zhang et al. (2020b) adopted a wider focus and examined the roles of different transport modes (air as well as high speed train and coach) in the spread of COVID-19 pandemic across Chinese cities.

Important Section:
The World Health Organization (WHO) declared the COVID-19 outbreak, with its epicentre in Wuhan, China, a Public Health Emergency of International Concern (PHEIC) (Lau et al., 2020b). Consequently, the Hubei province was placed under lockdown approximately three weeks after the start of COVID-19 outbreak. Lau et al. (2020a) compared domestic and international passenger volumes and routes of China to the distribution of domestic and international COVID-19 cases and found that there is a strong linear correlation within China and a significant correlation between international COVID-19 cases and passenger volumes. Cooper et al. (2006) and Keogh-Brown and Smith (2008) suggest that border closures are an expensive and ineffective outbreak control measure. With low vaccination rates and new virus variants, social distancing, which includes quarantine and travel restrictions is the main tool used. Some countries reacted by completely and immediately shutting down their borders and restricting flights to and from affected areas and others took fewer measures, implementing a 14-day self-quarantine for incoming travellers.

Data usage of Lao: 3. Data
The airline analysis here is both demand and supply orientated, evaluating the Chinese domestic market, but also the China to Europe and China to other Asian destinations. The supply analysis was conducted using Official Airline Guide (OAG) data. OAG is a comprehensive subscription database that records 96% of global passenger itineraries. OAG has been used in various academic papers (e.g. Corbet et al., 2019; Lei and O'Connell, 2011; Pagliari and Graham, 2019). This database does not include charter flights and in this analysis, cargo flights are not included. Daily supply data reported by origin-destination (O-D) pairs from January 2017 to December 2020 was collected.

--

International air travel can accelerate the spread of infectious diseases. Hufnagel et al. (2004) studied the spread of SARS and found that the isolation of large cities can be an effective epidemic control measure, but Cooper et al. (2006) in their study on influenza found this does not apply equally to influenza that has a much shorter serial interval than SARS. Tuncer and Le (2014) used a two-city dispersal model to predict the spread of avian influenza from Asia and Australia to major US cities via air travel and found that the effectiveness of control measures (e.g. quarantine and vaccination) depends strongly on the air travel rate. Brownstein et al. (2006) found that a decrease in domestic and international air travel was associated with a delayed and prolonged influenza season. Cooper et al. (2006) suggest that under most scenarios of air travel restrictions that they tested there is little value in delaying epidemics, unless almost all travel ceases very soon after epidemics are detected. Grais et al. (2003) suggest that the time lag for public health intervention is very short and coordinated pandemic planning is vital.
Cooper et al. (2006) argue that the rapid initial rate of growth of the epidemic in each city and the large number of people infected can make travel restrictions a relatively ineffective control measure. Lau et al. (2020b) argue that while the COVID-19 spread could not be contained, the lockdown in Hubei aided in slowing the speed of infection and reduced the correlation of domestic air traffic with COVID-19 cases within China. Lau et al. (2020b) believe that it is not feasible to contain the global spread of COVID-19, especially when considering political willingness and feasibility constraints to implement drastic countermeasures with tremendous social and economic consequences.


Zeng et al. (2005) suggest that tourism exhibits little resistance, but considerable resilience. Similarly, air transport is hugely impacted by shocks, disasters and epidemics, but its recovery period and type of effects differs depending on the nature, intensity and duration of the shock (Corbet et al., 2019). During the SARS epidemic of 2003, traffic dropped significantly. IATA Economics (2020a) suggests that the SARS epidemic had a serious impact on traffic volumes for Asia/Pacific based carries (Fig. 1). In 2003, Asia-Pacific airlines lost US$6 billion after seeing a reduction of 8% in Revenue Passenger Kilometres (RPKs), and 35% reduction at the peak of outbreak in May 2003.

Warnock-Smith et al about covid 19 and chinese new year
2.3. COVID-19
COVID-19 was initially recorded as an unidentified coronavirus in Wuhan, China at the end of December 2019 and was classified as the COVID-19 pandemic on March 11, 2020 by the World Health Organisation (WHO). The Chinese market has noticed very strong growth the recent years with an additional 450 million passengers per year flying to/from and within China compared with a decade ago (IATA Economics, 2020a). The COVID-19 outbreak coincided with the busiest travel season for China, i.e. the Chinese New Year. Frequencies of flights and high-speed rail services out of Wuhan related to the number of COVID-19 cases in the destinations (Zhang et al., 2020b). Therefore, a number of travel restrictions were imposed on Chinese travellers leading to the collapsing of traffic. China was the first to experience the pandemic and has now entered a ‘restart phase’. COVID-19 is an unprecedented situation having significant impact on flight volumes for a prolonged period (Dube et al., 2021). According to UNWTO (2020) in May 2020 from 217 destinations worldwide, 97 (45%) of them partially or completely closed their borders to tourists, 65 destinations (30%) suspended international flights partially or completely and 39 destinations (18%) banned the entry for passengers from specific countries of origin or passengers who have transited through specific destinations (Fig. 2). When the COVID-19 situation escalated, further travel restrictions were imposed (Fig. 3).

Iacus et al. (2020) used trends analysis to reflect the impact of reduced traffic demand and lockdowns in Europe and found that international flights decreased significantly. They also highlighted that different lockdown strategies impact aviation in different ways. Budd et al. (2020) found that the pandemic resulted in a contraction of fleet size, labour and network coverage. Carriers reported 50% no-shows for flights and reduced future bookings (IATA, 2020a, IATA, 2020b, IATA, 2020ac, IATA, 2020bc). The industry urged regulators to revise exiting rules and regulations and show flexibility. According to Sun et al. (2020), the global airport network remained unchanged for the first 2–3 months. This can be explained by the slot regulation in congested airports. With 43% of global passenger traffic departing from over 200 slot coordinated airports (IATA, 2020b), airlines initially flew ghost flights to comply with slot rules. These rules were relaxed initially for operations to China and Hong Kong, and after industry requests, the EC revised the EU Airport Slots Regulation (EEC 95/93). Consequently, capacity fell even further. The aviation fee reductions and cost support that many countries, including China, offer, contributes to the reduction of airlines' marginal costs, but are not sufficient to make the carriers profitable according to Czerny et al. (2020).


Warnock-Smith also using OAG just as Lao et al. (2020) did (! Cargo flights not included) :

3. Data
The airline analysis here is both demand and supply orientated, evaluating the Chinese domestic market, but also the China to Europe and China to other Asian destinations. The supply analysis was conducted using Official Airline Guide (OAG) data. OAG is a comprehensive subscription database that records 96% of global passenger itineraries. OAG has been used in various academic papers (e.g. Corbet et al., 2019; Lei and O'Connell, 2011; Pagliari and Graham, 2019). This database does not include charter flights and in this analysis, cargo flights are not included. Daily supply data reported by origin-destination (O-D) pairs from January 2017 to December 2020 was collected.

...

For the airport analysis, the OAG data for the domestic, Asian, European and North American markets was used to track overall trends at the airports before and after COVID-19. Whilst this does not capture all activity at the airports, it includes all the main markets and so was considered a fair representation of the situation.



Do i see the same pattern -> Europe?

4.1. The overall picture
The overall picure in China in terms of air carrier seat capacity, passenger traffic and passenger revenues has been mixed depending on the market, though in all cases there has been prolongued negative growth during 2020 in comparison with previous years. Fig. 5, Fig. 6, Fig. 7 show how seat capacity has developed over a 48 month time-series from January 2017 to December 2020 on China Domestic, China to Europe and China to rest of Asia markets respectively. The observed drops in capacity might have been even worse if the Chinese government had not introduced a payment scheme of US$0.0027 per ASK for flights on routes served by multiple airlines, and US$0.0081 per ASK for a route where the carrier was a sole operator to encourage the airlines to keep flying (Flightglobal, 2020). The sustained presence of the virus along with government imposed travel restrictions, as shown, for example, on China to Europe markets (Fig. 6) by the sudden drop in capacity in February 2020 and the absence of any recovery ever since on, made the opening up of international markets particularly difficult. Only the Chinese domestic market saw any pronounced rebound in capacity. Europe and rest of Asia markets continued to languish at 7% and 8% of 2019 capacity levels respectively by December 2020 (Fig. 7).

Aside from the normal seasonal variation, which can be seen on the more leisure intensive China to Europe markets in comparison to the China Domestic and China to rest of Asia markets, year-on-year increases in capacity up to the end of 2019 can be observed, followed by significant capacity drops as lockdown measures and travel restrictions took hold first in China in January 2020, and then by other countries from February 2020 onwards. As early as April 2020, China Domestic capacity had partially rebounded with levels being as much as 63% of those observed in April 2019. By December 2020, domestic seat capacity had recovered to 93% of 2019 levels. A similar pattern can be observed in Fig. 8, Fig. 9, Fig. 10 for passenger traffic and passenger revenues on all three observed China traffic markets (Domestic, Europe and rest of Asia). In all cases by September 2020, passenger traffic and passenger revenues were smaller than September of the previous year and with the exception of the domestic China market (revenues and seats only), smaller than September 2017 levels as well.

There is some variation between markets that is worthy of note. China to Europe passenger revenues were already on a downward trend in 2019 before the virus hit, primarily linked to downward pressure on average fares on what were becoming increasingly competitive, lower yield markets. This was driven by large capacity increases by Chinese based carriers into European markets with Air China, China Eastern, China Southern and Hainan Airlines all seeing double digit growth in 2018 versus 2017, for example (Anna.aero, 2019). Responding to a slowing rate of GDP growth in China,1 airlines on these markets had to reduce fares in order to preserve the same number of passengers as shown particularly in Fig. 10. The Chinese domestic market (Fig. 8) saw year-on-year increases both in passengers revenues and passenger traffic despite small average fare increases through to September 2019, which is indicative of the more inelastic, expanding domestic market within China and the more controlled competitive environment in which air carriers operate. Remarkably, by September 2020, passenger traffic and revenues had rebounded and were only marginally below September 2019 levels, and the same or greater than September 2018 and 2017 levels.

--

The Chinese domestic market is quite dispersed given that the top 10 routes (inclusive of Hong Kong and Macao SARs) represent around 7% of the total 373 mn Chinese domestic seats available in 2019 (See Appendix A). Despite year-on-year growth in the overall supply (as shown in Fig. 5), all of the top 10 routes as outlined in Fig. 11 witnessed downturns in traffic, capacity and revenues from the beginning of 2020 onwards. By March 2020 the worst affected routes appear to be those involving Hong Kong (HKG) and to a lesser extent those involving Beijing (PEK). The traffic losses at these two airports overall was particularly severe (see section 4.5 for more detail). The HKG-PEK route, which was the 9th busiest route in 2019 (by capacity), appears to have been the worst affected with seats, passenger traffic and passenger revenues on the route being a mere 13%, 12% and 8% of 2017 levels in March 2020. In contrast, the least affected route appears to be CTU-PEK, the 5th busiest route in 2019 with market indicators partially rebounding in March 2020 from lows in the previous month. Capacity, passenger traffic and passenger revenues were back up to 48%, 58% and 110% of March 2017 levels respectively. By September 2020, market indicators on Hong Kong (HKG) routes had deteriorated even further with −96.5% year-on-year reductions in revenue on the HKG-PEK route for example, versus −89% in March (YoY). On the Hong Kong (HKG) to Shanghai Pudong (PVG) route, year-on-year traffic picked up only sightly in September 2020 versus March with revenues only 12% of September 2019 levels (up from 6% in March 2020 YoY). Conversely, market indicators on top 10 routes involving Beijing (PEK) appeared to have partially recovered by September 2020 with the biggest recovery being observed on the Beijing (PEK) to Shanghai Honqiao (SHA) route. Despite passenger traffic recovering to 78% of 2019 levels in September (from a base of 18% of 2019 levels in March), total revenues on the route were miraculously 1% higher than September 2019 levels, boyed by higher average fares ($220 in September 2020 versus $172 in September of the previous year).


Europe:

4.3. China to Europe route, airline level analysis and premium/economy revenue comparison
China to Europe is a more condensed and focussed network in comparison with the Chinese domestic market. The top 10 routes sorted by 2019 seat capacity represent 59% of all China to Europe seats in that year (Appendix A). HKG-LHR is by far the densent route followed by the Beijing and Shanghai to Moscow routes and then Beijing to other major European points including Frankfurt, Paris, Amsterdam and Munich. In part, due to the Far East network strategy focus of Finnair, owing to Helsinki's geographical location reducing Europe to China distances, HKG-HEL also makes it into the top 10 routes with a proportion of this traffic making connections from other European/Asian points.
Unlike the situation on the Chinese domestic market, prior to the COVID-19 crisis traffic growth had remained steady rather than bullish in the China to European market. There was already some evidence of reductions in average fares and revenues on these markets up to 2019. The COVID-19 pandemic excaerbated the revenue losses significantly and led to an unprecedended downturn in traffic after remaining steady in previous years. Fig. 15 provides a detailed comparison of trends in seats, average fares, revenues and passengers on the top 10 China to Europe routes in both March and September (over the 2017–2020 period).
Fig. 15

Due to the time difference in the initial spread of COVID-19, it can generally be observed that significant revenue drops in China to Europe markets did not take place until March 2020 at which point market indicators started to drop quite dramatically. The worst affected routes in March 2020 appeared to involve secondary gateways. Both PEK-MUC and HKG-MUC saw March 2020 revenue drop to as little of 2% of March 2017 levels (HKG-MUC). In the case of PEK-MUC there was still some remnance of service in March 2020, though with revenues down to as little as 9% of March 2017 levels on the back of seat capacity reductions to 32% of March 2017 levels. With already reducing infection rates in China during March 2020 and still controlled rates of infection in Russia, March 2020 saw PEK-SVO in particular being one of the least affected routes with traffic, seats and revenues all retaining between 50% and 60% of 2019 levels. Within this group of observed routes, HKG-LHR continued operations albeit at a reduced level. Interestingly, average fares spiked in March 2020 at US$1,052 up from US$923 in March 2019 and US$772 the previous month (Feb 2020). This reflects attempts by carriers on this route to try and compensate for falling load factors by charging non-discretionary, inelastic passengers higher fares, who would often consider their journeys to be essential during this period. By September 2020, the situation had actually worsened as international markets continued to bear the brunt of inconsistent government approaches to travel restrictions and COVID-19 response initiatives. With COVID-19 also having a big impact in Russia, top 10 routes like PEK-SVO that were looking a bit more resilient in March 2020 had also tapered off to negligible levels by September 2020. In the whole of September 2020, there were just 284 O&D passengers travelling on this route compared to over 13,000 in September of the previous year. Using the observed pairs, the situation by September 2020 was almost universally bleak. Only HKG-LHR had some remnants of traffic (13,000 versus 38,000 the year before), with average fares reducing to a low of US$713. Despite the mandatory quarantines that were in place in the UK at this time, the country was still partially open to international flights, including from China (and Hong Kong).
In terms of top 20 airlines serving China to Europe markets, despite not having any direct services, Emirates is a large player in this market, transporting passengers between Europe and China via its Dubai, UAE hub (Appendix B). Aside from Emirates, Air China, Aeroflot and Cathay Pacific are the largest players with some of the main European carriers such as British Airways, Lufthansa and Air France also in the top 10.

5. Conclusions
The aim of this paper was to analyse the impact of COVID-19 on Chinese air passenger markets through time by considering airline seats offered and passengers flown; airline revenues and average air fares; and airport frequencies. Overall it was found that those routes served by well financed/funded air carriers, those exposed to the lowest rates of COVID-19 infection and/or those that are seeing the least restrictive lockdowns and travel measures have been impacted least by the pandemic and are those that are most likely to rebound first.
It is very clear that not all air carriers have been impacted equally by the COVID-19 pandemic. Less well-financed/funded carriers whose networks are focussed on international markets, premium traffic and discretionary leisure travel have been found to be impacted most by the pandemic and are those that are likely to take the longest to recover. Conversely, better financed/funded airlines with a greater focus on domestic markets, non-discretionary traffic, and standard economy class fares have been found to be less severely impacted by the pandemic. Spring Airlines’ nimbleness as a large Shanghai-based LCC and its focus on domestic and intra-regional markets, has allowed it to rebound quicker and perform better than its peers during the observed period, even on the generally more restricted China to rest of Asia international markets.
In terms of Chinese airports, performance has varied according to airlines served, characteristics of the airport/city, and the severity of the outbreak. The experience of Chengdu has been contrasted with Wuhan for example, as has Hong Kong's with that of Guangzhou and Shenzhen. Reductions in traffic has caused very significant decreases in airport revenues and profits, especially for airports with large international traffic volumes such as those serving Beijing, Shanghai and Hong Kong. However, airport construction and capacity expansion have returned, indicating the government's optimism about the future.
As other regions and countries go through the different stages of their COVID-19 epidemics, this study has unearthed some useful indications from the Chinese experience for airports, airlines and route markets worldwide. Airports handling domestic traffic may fair better in the short term in countries where there is a lot of domestic traffic. Equally more diversified air carriers and route markets in terms of market segments served (i.e. those less reliant on international/intercontinental traffic) are also more likely to be less severely impacted by the pandemic and more likely to recover faster to pre-COVID-19 traffic and revenue levels.
This research has contributed to the body of knowledge by providing disaggregated evidence on the impact of COVID-19 on capacity and demand, including fares and revenue. The paper has linked the traffic volume and frequency to industry responses, mainly those of airlines and airports. One of the most significant global traffic markets (in ASK and RPK) has been investigated and most importantly this is the market from where the pandemic started. The research provides evidence on the scale of the disruption, but also on the possible pathways to recovery. It also opens the door for further research. For example, at a later stage, once the situation appears more stable, a causality analysis could be undertaken with the data to provide useful and future insight into the key factors that drove traffic patterns during the pandemic.
A primary limitation of this research is that it has not considered air cargo traffic despite the fact that it has been less severely affected by the COVID-19 pandemic than passenger markets. It would be important for any follow-up research to consider whether combination carriers and route markets that have a more diversified mix of traffic will be better placed to deal with high impact external events such as a pandemic. There are also limitations in terms of the data. Sabre passenger demand, fare and revenue data was only available up to September 2020 at the time of this research. It was not possible therefore, particularly for the China to Europe market to observe the year 2020 as a whole, and how second waves during colder winter periods, particularly in Europe further hindered the air traffic recovery.



* **Two-month window (Dec 2019 & Mar 2020).**  
  EUROCONTROL’s public archive releases only four sample months per year; February—a critical seeding month—is missing [@eurocontrolData2025].

* **Excess-mortality uncertainty.**  
  Outcome data merge the World Mortality Dataset with HMD-STMF. National registration lags, baseline-fit choices, and wartime disruption (Ukraine ≥ 2022) widen credible intervals [@karlinsky2021; @msemburi2023]. Notably, @msemburi2023. caution that "basing rankings on point estimates will often be misleading" and discuss the "hazards of ranking countries' epidemic responses" given large data and model uncertainties.
  – A ± 7-day tolerance around 5 May grabs the nearest non-NA value (day_diff ≤ 7); this reduces missingness but adds timing noise.
  – Country ranks are more stable than absolute levels, yet still volatile for very small populations.
  – As emphasized by @msemburi2023, "hazards of ranking countries’ epidemic responses based solely on excess-mortality estimates should be acknowledged."

* **Population denominator.**  
  OWID’s mid-2020 snapshot (UN WPP 2024–derived) differs from raw WPP by < 5 % in 38 / 41 countries; the largest gaps (Cyprus –31 %, Ukraine –11 %, North Macedonia +11 %) stem from boundary or census updates. Re-running all correlations with WPP figures shifts Spearman ρ by **< 0.01**, so OWID values were retained for consistency.

### Methodological  

* **Ecological, country-level design.**  
  Results are based on national aggregates, which mask substantial sub-national variation in both air-traffic exposure and health outcomes (e.g. Lombardy vs. Sicily in Italy).
  – This approach risks the ecological fallacy: relationships observed at the country level may not hold at the city, regional, or individual level.
  – Therefore, findings should not be interpreted as causal or representative for specific regions or population subgroups.


* **Small-N, rank-based correlation.**  
  With ≈ 25 observations, Spearman ρ is sensitive to ties and outliers; one hub country can swing the estimate. We report 5 000-replicate bootstrap CIs, but sampling error remains large [@diciccio1996].


* **Residual confounding.**  
  Excess-mortality differences across countries are strongly shaped by vulnerability factors—GDP, poverty, and income inequality. We do **not** adjust for these, so findings may reflect socioeconomic risk more than direct effects of air connectivity [@ioannidis2023]. - Ward et al 2024 - msemburi et al 2023 ..
  


### Causal interpretation  

* **No causal identification.**  
  December exposure precedes the European outbreak (minimising reverse causation), but March exposure is already influenced by border closures, so causality could run either way.

* **External validity.**  
  Findings apply to continental EUROCONTROL members during the Wuhan/Alpha period. Extrapolation to later variants, island nations, or other regions should be cautious.

> **Bottom line:** The study is *hypothesis-generating*. Direct air connectivity plausibly seeded Europe’s first wave, but its statistical signal disappears once community transmission dominates. Data gaps, ecological design, and unmeasured confounders limit strong policy conclusions about long-run border-control effectiveness.

# Robustness check

```{r robustness_table}
boot_rho <- function(data, idx, xvar) {
  with(
    data[idx, ],
    cor(get(xvar),
        excess_mortality_cumulative_per_million,
        method = "spearman",
        use   = "complete.obs")
  )
}

# NEXT ! 

```

## Reverse causation

Early flight exposure is measured **before** first-wave mortality peaks; therefore, high deaths could not have caused higher December exposure. Correlation using **Dec-only** flights (ρ ≈ 0.50) shows the pattern predates policy cuts.

## Take-away  

> **Early seeding signal, rapidly eclipsed.**  
> Direct flights from China / Hong Kong explain roughly one-quarter of the *cross-country* spread in first-wave excess mortality (May 2020). That signal fades to zero in 2021 and reverses thereafter, once community transmission and domestic counter-measures dominate outcomes.


